---
title: "Incarceration Trends In the U.S. (1970-2018)"
author: "Allison Huang"
date: 'Autumn 2022'
output: html_document
---

```{r echo=FALSE, eval=TRUE, message=FALSE, include=FALSE}
source("../source/a4-helpers.R")
source("../source/analysis.R")
```

## Abstract 

**Incarceration rates** across the U.S. give *insight* to more
than just jail populations alone, but how they relate to **discriminatory
factors** and changing policies. The dataset points to *intersections*
between jail populations, and the backgrounds and **identities** of people affected.
Additionally, the data reveals trends over time, and how numbers are distributed
geographically.

```{r echo=FALSE, eval=TRUE, message=FALSE, include=FALSE}
source("../source/a4-helpers.R")
source("../source/analysis.R")
```
## Introduction

The incarceration rates across the U.S. over the years reveals a lot about
the country, including **social issues** and **inequalities**.
Specifically, the data exposes patterns of inequality, including those on
one's *identity and background*, such as age, gender, and race,
leading to social issues that have arisen as a result.
For example, according to data, Black people are much more likely to
be arrested in comparison to White people, suggesting
deeper problems within social systems and policies set in place
in communities (vera.org). Additionally, people of color and other minorities
are much more likely to be incarcerated as well too. Differences in
rates could mean those in power created policies that
give certain groups advantages and others disadvantages, especially those
that are already more vulnerable in society.  

**Direct stakeholders** include people living in the U.S.
in general. More specifically, the audience are people that are
incarcerated, and those in groups disproportionately put into jail,
so that it can be taken note of to fight for justice in existing inequalities.
**Indirect stakeholders** include journalists, policymakers, and government officials.
Journalists would use the data to report to the public
about incarceration rates in the United States. Policymakers and government
officials would refer to data to attempt to lessen inequalities that
exist simply due to people's backgrounds, and train officials so that
people are arrested equally due to the right reasons. With this, the
central values that the dataset needs to do is be informative and
clear, so that people of all levels can interpret the data, and so that the
data can be communicated to be used as power to fight against inequalities that occur.  

**Questions** that will be answered in the report relate to growth of jail
populations in the U.S., jail population by race, and jail populations
distributions across states.  
* How have the overall number of jail populations across the country
of the United States changed over time?  
* How does the jail population in different states vary and
change over time?  
* How does the jail population of people belonging to different races
compare, and what are the trends of those racial populations in jail over time?  
* How do jail populations across the U.S. in different states vary in
the year of 2018? 


## Data Summary

```{r echo=FALSE, eval=TRUE, message=FALSE, include=FALSE}
source("../source/a4-helpers.R")
source("../source/analysis.R")
```

```{r echo=FALSE, eval=TRUE, message=FALSE,}
```

Over the years, the change in the jail population has been in general quite
a steep increase. Specifically, within this dataset, there has been an giant
increase of about **`r change_1970_2008`** in total jail population from the years of
1970 to 2008. However, the change in jail population has been less in
comparison to that, within the most recent ten years of the data (2008-2018),
with a slight decrease of about **`r change_2008_2018`** in total jail population.
To add on, in the most recent year of 2018 in the dataset, the average jail
population by state across the U.S. was **`r avg_2018_jail_pop`**. With that,
**`r highest_state_jail_pop`** was the state with the highest jail population
at **`r highest_avg_jail_pop`**. Additionally, the dataset also reveals inequalities
in jail population. For example, over the years certain racial groups have made
up on average more of the total jail population than other races have. In this case,
the two groups with the highest average jail population per year would be
Black people with an average of **`r avg_black_jail_pop`**, making up
**`r percent_black_jail_pop`** of the total jail population, and White people with
an average of **`r avg_white_jail_pop`**, making up **`r percent_white_jail_pop`** of the total population.


## Growth of the U.S. Prison Population

```{r figure_1, echo=FALSE, eval=TRUE, message=FALSE, include=FALSE}
source("../source/a4-helpers.R")
source("../source/analysis.R")
```

``` {r plotting figure_1, echo = FALSE, warning = FALSE}
plot(figure_1)
```

Question: How have the overall number of jail populations across the country
of the United States changed over time?  
The bar chart depicting the U.S. jail population over the years is significant
as it points to a trend in which there has been great increases in jailings
across the country. The data is important as people can visually see that since
1970, the number of jailings have nearly quadrupled, which is quite alarming,
and could point to other factors that are causing this. Additionally, based on
the graph, the steepest increases occurred between the years 1980 and 2005.


## Growth of Prison Population by State

```{r figure_2, echo=FALSE, eval=TRUE, message=FALSE, include=FALSE}
source("../source/a4-helpers.R")
source("../source/analysis.R")
```

``` {r plotting figure_2, echo = FALSE, warning = FALSE}
plot(figure_2)
```

Question: How does the jail population in different states vary and
change over time?  
The line graph displays the number of people in jails by state over the time
period between 1970 and 2018. From the chart, it can be clearly seen that
the state of California has the highest jail populations
for all of the years that are included in the data, which could
be interesting, as it potentially points to the systems that are put into place
in the state of California and specific locations that could be further looked into.
In addition, both the states of CA and FL seem to have spikes and jumps in the
number of jailings between about 1978 and 2008, in which there were great increases
in their respective jail populations. Also, for all of the states that
are shown in the chart, they all are depicted to have increases in their state
jail populations. The states of AL, CA, FL, NV, NY, and WA were chosen
as many of these states are spread out in different sections of the United States.
Additionally, there are also a couple states chosen that are next to each other
geographically on the US map, for those that want to see how states that
are so close in vicinity could also have very different jail populations.


## Prison Population by Race Over Time Reveals Inequality

```{r figure_3, echo=FALSE, eval=TRUE, message=FALSE, include=FALSE}
source("../source/a4-helpers.R")
source("../source/analysis.R")
```

``` {r plotting figure_3, echo = FALSE, warning = FALSE}
plot(figure_3)
```

Question: How does the jail population of people belonging to different races
compare, and what are the trends of those racial populations in jail over time?  
The graph is significant because it shows groups of people of different
racial backgrounds and identities that are disproportionately affected by
the jail system, regardless of what the reason
may be. It also shows how those trends continue over time, from the 1900s to
the present day, therefore signifying that jailings according to race
may still be a great issue for politicians and researchers alike to look into.
It is interesting to see that over the years, regardless of what the total
jail population is, Black people and White people, always seem to have
the most people in jail. It is also notable that from the years of 1970
all the way up to nearly 1985, there is no data on the
bar chart, as data could not have been compared since the data for some of the
races were not available. This is also significant as the data being unavailable could possibly
point to the idea that data in regards to race had not been collected
at the time, which also points to how times may have potentially
changed and people started emphasizing the importance in the
disparities between race, which in this case, is the jailing system in the U.S. over time.


## Prison Population By State Reveals Inequality

```{r figure_4, echo=FALSE, eval=TRUE, message=FALSE, include=FALSE}
source("../source/a4-helpers.R")
source("../source/analysis.R")
```

``` {r plotting figure_4, echo = FALSE, warning = FALSE}
plot(figure_4)
```

Question: How do jail populations across the U.S. in different states vary in
the year of 2018?  
The graph is significant because it shows how different states across the U.S.
have different amounts of jail populations. It can be seen that many of the
states in the U.S. had similar numbers in jail populations due to the color
fill of navy that they consist of. However, areas that are more of a darker
red and magenta color had greater jail populations, such as the states of
New Mexico, Colorado, and Idaho. Furthermore, California has a brighter red,
and is also the color highest on the scale for this map, signifying that
they had the most jail population in the year of 2018. In addition, the year
of 2018 was chosen as it is the most recent to the present day in regards to
this dataset. With such existing variances, it would be interesting for
people to dive into the reasonings of why such inequalities occur, and if
they point to differing policies in the different states.

